A Formal Model of Checked C

### 2023-01-19 sandp23/introduction.tex

Vulnerabilities due to memory corruption, especially spatial memory corruption, 
are still a major issue for C programs~\cite{cvetrend, microsoftmemsafe, Zeng:2013:SRF:2534766.2534798} 
despite many efforts that tried to prevent them~\cite{song2019sanitizing}.
Several industrial and research efforts, including CCured~\cite{Necula2005},
Softbound~\cite{softbound}, and ASAN~\cite{Serebryany2012},
have investigated ways to better compile C programs with automatic spatial safety enforcement.
These approaches all impose performance overheads deemed too high for deployment use. 
% 
%\liyi{Should we say this? Is the paper about overhead reducing? }
Recently,~\citet{Elliott2018} and~\citet{li22checkedc} introduced and formalized \checkedc, an
open-source extension to C,
to ensure a program’s spatial safety by introducing new pointer types,~\ie checked pointer types.
%
The checked pointers are represented as system-level memory words without ``fattening'' metadata~\cite{duck2016heap}, 
and ensuring backward compatibility,~\ie developers can use checked and regular (unchecked) pointers with the same program.
%users are able to split code into checked and unchecked regions and incrementally convert
%C code in the unchecked region to Checked C code in the checked one.
However, as we explain in~\sect{subsec:nosafetyagsintuncheckedcode}, the
unconverted or unchecked code can violate guarantees provided in checked
regions.
%
We need to ensure that~\emph{code executed as part of 
unchecked regions does not violate the safety guarantees in checked regions  with the use of program partitioning mechanism \cite{rul2009towards}}.

@@

Existing such mechanisms are not suitable as they are based on process isolation and have high overhead, and are \emph{data-centric} (\sect{subsec:background:programpart}).
%Furthermore, these techniques are hard to engineer to co-exist with~\checkedc.
%\liyi{why? Can we not say this? It seems that the whole reason people should care our work is we did it in \checkedc, but \checkedc itself is the selling product here.... }
But in our case, we want a low-overhead code-centric partitioning, where the unchecked code (or functions) should be isolated (or partitioned) from checked code. We also want the technique to co-exist and be compatible with Checked C guarantees such that the partition containing checked code should still enjoy its spatial safety.

@@

The developer starts by marking desired (\ie unchecked) functions and pointers used in functions as tainted. 
Then, \systemname partitions the given program into two partitions (\umode and~\cmode regions)  of different privileges:

\begin{itemize}
\item~\umode \emph{region} or~\emph{tainted region} (low privilege tainted region): This partition contains tainted types (\ie functions and pointers) and can only access tainted pointers.
\item~\cmode \emph{region} or~\emph{safe region} (high privilege untainted or checked region): This partition contains the remaining (untainted) code and data and has complete access to~\cmode region.
The functions in~\cregion can invoke any function in~\ucregion and access all its data but not the other way around, except for call-back functions, which we will discuss later.
\end{itemize}

The $\cmode$ region code is executed as a regular program, while the~\ucregion partition will be executed in an existing sandboxed environment (\eg WASM sandbox), and our compiler adds necessary instrumentations to facilitate the communication between code in~$\cmode$ and~$\umode$ regions.

The combination of tainted types and privileged partitions 
enables us to enforce isolation and provide memory safety without transforming all unchecked C code to \checkedc code,
because unchecked types can stay in \ucregion, and \cregion code can access tainted type entities that are allocated 
in \ucregion.
% 
Although memory isolation prevents direct violations,~\ucregion code can still affect~\cregion through tainted pointers by confused deputy attacks~\cite{rajani2016access, machiry2017boomerang},~\eg by using a valid~\cregion address in a tainted pointer.
Our compiler avoids these attacks by ensuring using dynamic checks that tainted pointers validly point to~\ucregion address space.
Such checks are statically generated by our compiler.

@@

\myparagraph{\systemname Type System, Formalism and Compiler}
We present a type system that integrates tainted types with~\checkedc and provides additional guarentees,~\ie the~\emph{non-crashing} and \emph{non-exposure} guarantees,~\ie a well-typed~\systemname program can never crash due to spatial safety violations,
as well as \ucregion code cannot directly observe a checked pointer address.
%no \cregion pointer addresses will be leaked in
%\ucregion code
We extend~\checkedc compiler to support our type system.
We formalize our system by extending~\checkedc formalism~\citet{li22checkedc} with the non-crashing and non-exposure guarantees.
We formally prove theorems related to the two guarantees and use model-based randomized testing (CITE) to certify the simulation relation between the~\systemname semantics and its compiler formalism.
To the best of our knowledge,~\systemname is the first C(-like) language and compiler formalism with the program partitioning mechanism.

\myparagraph{Type-Directed Program Partitioning}
We present a type-directed program partition technique to separate $\cmode$ and $\umode$ code regions and ensure the above guarantees.
%Our type system restricts checked pointers usage only in~\cregion and ensures that~\ucregion can only access tainted pointers.
Our modular design enables us to use existing sandboxed techniques to enforce memory and execution isolation, with the implementation of tainted pointers in the \systemname compiler.

\myparagraph{Supporting callbacks to~\cregion with no Checked Pointer Exposure}
Although we disallow access to~\cregion from~\ucregion, there can be cases where such access is needed.
Specifically, when~\cregion wants to provide access to certain shared checked data to~\ucregion.
To enable this, we support callback functions in~\cregion that can be invoked from~\ucregion through function pointers.
%However, knowing the address of~\cregion functions in~\ucregion violates the program partition principle of \systemname.
However, knowing the address of~\cregion functions in~\ucregion violates the non-exposure guarantee and leads to other attacks~\cite{hauser2019sleak}.

>>>>

Vulnerabilities due to memory corruption, especially spatial memory corruption, 
are still a major issue for C programs~\cite{cvetrend, microsoftmemsafe, Zeng:2013:SRF:2534766.2534798} 
despite many efforts that tried to prevent them~\cite{song2019sanitizing}.
Several industrial and research efforts, including CCured~\cite{Necula2005},
Softbound~\cite{softbound}, and ASAN~\cite{Serebryany2012},
have investigated ways to better compile C programs with automatic spatial safety enforcement.
These approaches all impose performance overheads deemed too high for deployment use. 
% 
%\liyi{Should we say this? Is the paper about overhead reducing? }
Recently,~\citet{Elliott2018} and~\citet{li22checkedc} introduced and formalized \checkedc, an
open-source extension to C,
to ensure a program’s spatial safety by introducing new pointer types,~\ie checked ($\cmode$) pointer types.
%
The checked pointers are represented as system-level memory words without ``fattening'' metadata~\cite{duck2016heap}, 
and ensuring backward compatibility,~\ie developers can use checked and regular (unchecked $\umode$) pointers within the same program.
%users are able to split code into checked and unchecked regions and incrementally convert
%C code in the unchecked region to Checked C code in the checked one.
However, as we explain in~\sect{subsec:nosafetyagsintuncheckedcode}, the
unconverted or unchecked ($\umode$) code can violate guarantees provided in $\cmode$
regions.
%
We need to ensure that~\emph{code executed as part of 
unchecked ($\umode$) regions does not lead to the safety violations in checked ($\cmode$) regions  with the use of program partitioning mechanism \cite{rul2009towards}}.

@@

Existing such mechanisms are not suitable as they are based on process isolation and have high overhead, and are \emph{data-centric} (\sect{subsec:background:programpart}).
%Furthermore, these techniques are hard to engineer to co-exist with~\checkedc.
%\liyi{why? Can we not say this? It seems that the whole reason people should care our work is we did it in \checkedc, but \checkedc itself is the selling product here.... }
But in our case, we want a low-overhead code-centric partitioning, where the $\umode$ region code (or functions) should be isolated (or partitioned) from $\cmode$ one. We also want the technique to co-exist and be compatible with Checked C guarantees such that the partition containing $\cmode$ region code can maintain spatial safety.

@@

The developer starts by marking desired (\ie unchecked, $\umode$) functions and pointers used in functions as tainted. 
Then, \systemname partitions the given program into two partitions (\umode and~\cmode regions) of different privileges:

\begin{itemize}
\item~\umode \emph{region} (low privilege tainted region, extended from the unchecked region in \checkedc): this partition contains tainted types (\ie functions and pointers) and can only access tainted and unchecked pointers.
\item~\cmode \emph{region} or~\emph{safe region} (high privilege untainted or checked region): This partition contains the remaining (untainted) code and data and has complete access to~\cmode region.
The functions in~\cregion can invoke any function in~\ucregion and access all its data but not the other way around, except for call-back functions, which we will discuss later.
\end{itemize}

The $\cmode$ region code is executed as a regular program, while the~\ucregion partition will be executed in an existing sandboxed environment (\eg WASM sandbox), with additional instrumentations to facilitate the communication between code in~$\cmode$ and~$\umode$ regions.

The combination of tainted types and privileged partitions
enforces isolation and provides memory safety without transforming all unchecked C code to \checkedc code,
because unchecked types can stay in \ucregion, and \cregion code can access tainted type entities that are allocated 
in \ucregion.
% 
Although memory isolation prevents direct violations,~\ucregion code can still affect~\cregion through tainted pointers by confused deputy attacks~\cite{rajani2016access, machiry2017boomerang},~\eg by using a valid~\cregion address in a tainted pointer.
Our compiler avoids these attacks by ensuring using dynamic checks that tainted pointers validly point to~\ucregion address space.
Such checks are statically generated by our compiler.

@@

\myparagraph{\systemname Type System, Formalism and Compiler}
We present a type system that integrates tainted types with~\checkedc and provides additional guarentees---the~\emph{non-crashing} and \emph{non-exposure} guarantees,~\ie a well-typed~\systemname program can never crash due to spatial safety violations,
as well as \ucregion code cannot directly observe a checked pointer address.
%no \cregion pointer addresses will be leaked in
%\ucregion code
We extend the \checkedc compiler to support the type system and
formalize it by extending~\checkedc formalism~\citet{li22checkedc} with the non-crashing and non-exposure guarantees.
We formally prove theorems related to the two guarantees and use model-based randomized testing (CITE) to certify the simulation relation between the~\systemname semantics and its compiler formalism.
To the best of our knowledge,~\systemname is the first C(-like) language and compiler formalism with the program partitioning mechanism.

\myparagraph{Type-Directed Program Partitioning}
We present a type-directed program partition technique to separate $\cmode$ and $\umode$ code regions and ensure the above guarantees.
%Our type system restricts checked pointers usage only in~\cregion and ensures that~\ucregion can only access tainted pointers.
Our modular design enables us to use existing sandbox techniques to enforce memory and execution isolation, with the implementation of tainted pointers in the \systemname compiler.

\myparagraph{Supporting callbacks to~\cregion with no Checked Pointer Exposure}
Although we disallow access to~\cregion from~\ucregion directly, there can be cases where such access is needed.
Specifically, when~\cregion wants to provide access to certain shared checked data to~\ucregion.
To enable this, we support callback functions in~\cregion that can be invoked from~\ucregion through function pointers.
%However, knowing the address of~\cregion functions in~\ucregion violates the program partition principle of \systemname.
However, knowing the address of~\cregion functions in~\ucregion violates the non-exposure guarantee and leads to other attacks~\cite{hauser2019sleak}.
